---
title: '**DSC-680-Z1 Research Practicum** <br> Exploratory Data Analysis'
output:
  html_document:
    highlight: null
    mathjax: null
    prettydoc::html_pretty: default
    theme: null
  pdf_document: default
  word_document: default
---
## **Project Description**
The research practicum involves on-site experiential learning in a research setting. This setting may be in the private or public sector, it may include such locations as education, governmental, non-governmental, or general
research organization. The experience must provide students the opportunity to collect and analyze data, consider ethical implications of research, and draw empirically grounded conclusions.

<b>Purpose:</b><br>
Carry out exploratory data analysis on a set of random sample data extracted for machine learning.<br>
<b>University Name:</b> Utica College <br>
<b>Course Name:</b> DSC-680-Z1 Research Practicum <br>
<b>Student Name:</b> Henry J. Hu <br>
<b>Program Director Name:</b> Dr. McCarthy, Michael <br>
<b>Runtime Environment:</b> RStudio<br>
<b>Programming Language:</b> R <br>
<b>Original Data Frame:</b>
12,705,553 international wires belonging to 139 customers from 3 continents for the entire year of 2020.<br>
<b> Last Update:</b> July 21st, 2021

## **Clearing R Studio Memory Usage**

```{r, echo=TRUE, message=FALSE}
gc()
rm(list = ls())
```

## **Time Counter Start**
```{r, echo=TRUE, message=FALSE}
start_time <- Sys.time()
```

## **Include the knitr package for integration of R code into Markdown**

```{r, echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## **All the libraries used in this code**
```{r, echo=TRUE, message=FALSE}
library(readr)
library(easypackages)
libraries("caret","caretEnsemble","caTools","class","cluster","data.tree","devtools","doSNOW","dplyr","e1071","factoextra","gbm","FNN","FSelector","ggalt","ggforce","ggfortify","ggplot2","gmodels","klaR","lattice","mlbench","modeest","nnet","neuralnet","outliers","parallel","psych","purrr","readr","rpart","rpart.plot","spatialEco","stats","tidyr","randomForest","ROSE","rsample","ROCR","pROC","glmnet","gridExtra","R6","Epi") 
```

## **Import data into RStudio**

```{r, echo=TRUE, message=FALSE}
# input_data <- read_delim("Final_cleaned_data.txt", ",", escape_double = FALSE, col_types = cols(
#               TRANSACTION_ID = col_character(),
#               TRANSACTION_TIME = col_datetime(),
#               TRXN_MONTH = col_character(),
#               CLIENT_ID = col_character(),
#               COUNTRY_NAME = col_character(),
#               COUNTRY_CODE = col_character(),
#               CONTINENT_NAME = col_character(),
#               CONTINENT_CODE = col_character(),
#               SWIFT_MSG_TYPE = col_character(),
#               AVG_TRXN_AMT = col_double(),
#               TRANSACTION_AMOUNT = col_double()
#               ),
#     trim_ws = TRUE)
```


## **Data Sampling**

```{r, echo=TRUE, message=FALSE}

# Set random seed
set.seed(42)

# Sample the data
# input_data_4M <- input_data[sample(nrow(input_data), 4000000), ]

# Write data to storage
# write.table(input_data_4M, file="sample_df_4M.txt", append = FALSE, sep = ",", dec = ".", row.names = FALSE, col.names = TRUE)
# write.csv(input_data_4M,"sample_df_4M.txt", row.names = FALSE)

# Load data into data frame
input_data_eda <- read_delim("sample_df_4M.txt", ",", escape_double = FALSE, col_types = cols(
                TRANSACTION_ID = col_character(),
                TRANSACTION_TIME = col_datetime(),
                TRXN_MONTH = col_integer(),
                CLIENT_ID = col_character(),
                COUNTRY_NAME = col_character(),
                COUNTRY_CODE = col_character(),
                CONTINENT_NAME = col_character(),
                CONTINENT_CODE = col_character(),
                SWIFT_MSG_TYPE = col_character(),
                AVG_TRXN_AMT = col_double(),
                TRANSACTION_AMOUNT = col_double()
              ),
    trim_ws = TRUE)
```

## **Numeric/character field separator**

```{r, echo=TRUE, message=FALSE}
num.names <- input_data_eda %>% select_if(is.numeric) %>% colnames()
ch.names <- input_data_eda %>% select_if(is.character) %>% colnames()
```

## **Descriptive Statistics** 

These descriptive statistics reveal both the central tendency and dispersion tendency of the sample data for machine learning.

### **Dimension of data frame**
```{r, echo=TRUE, message=FALSE}
dim(input_data_eda)
```
### **Structure of data frame** 
```{r, echo=TRUE, message=FALSE}
str(input_data_eda)
```
### **Summary statistics of data frame**
```{r, echo=TRUE, message=FALSE}
summary(input_data_eda)
```
### **Mode of each variable**
```{r, echo=TRUE, message=FALSE}
lapply(input_data_eda[,num.names],mfv)
```
### **Variance of each variable**
```{r, echo=TRUE, message=FALSE}
lapply(input_data_eda[,num.names],var)
```
### **Standard deviation of each variable**
```{r, echo=TRUE, message=FALSE}
lapply(input_data_eda[,num.names],sd)
```
### **Glimpse of data frame** 
```{r, echo=TRUE, message=FALSE}
glimpse(input_data_eda)
```
### **Head of data frame** 
```{r, echo=TRUE, message=FALSE}
head(input_data_eda)
```
### **Tail of data frame**
```{r, echo=TRUE, message=FALSE}
tail(input_data_eda)
```

## **Correlation plot**
```{r p1, echo=TRUE, message=FALSE}
oldw <- getOption("warn")
options(warn = -1)
pairs.panels(input_data_eda[,num.names],gap=0,bg=c("green","red","yellow","blue","pink","purple"),pch=21)
options(warn = oldw)
```

## **Segregate and prepare data for bar plots** 

```{r, echo=TRUE, message=FALSE}

input_data_eda$AVG_TRXN_AMT=input_data_eda$AVG_TRXN_AMT/1000

input_data_eda <- input_data_eda %>%
  mutate(
          MONTH_TEXT = 
           case_when(
                      TRXN_MONTH == 1 ~ "Jan",
                      TRXN_MONTH == 2 ~ "Feb",
                      TRXN_MONTH == 3 ~ "Mar",
                      TRXN_MONTH == 4 ~ "Apr",
                      TRXN_MONTH == 5 ~ "May",
                      TRXN_MONTH == 6 ~ "Jun",
                      TRXN_MONTH == 7 ~ "Jul",
                      TRXN_MONTH == 8 ~ "Aug",
                      TRXN_MONTH == 9 ~ "Sep",
                      TRXN_MONTH == 10 ~ "Oct",
                      TRXN_MONTH == 11 ~ "Nov",
                      TRXN_MONTH == 12 ~ "Dec"
                    )
         )

NN_103_df <- input_data_eda[input_data_eda$CONTINENT_CODE	=='NN' & input_data_eda$SWIFT_MSG_TYPE=='103',]
NN_103_df <- NN_103_df[,c(3,10,12)]
NN_103_df = NN_103_df %>% distinct()
NN_103_df <- NN_103_df[order(NN_103_df$TRXN_MONTH),]
glimpse(NN_103_df)
dim(NN_103_df)

NN_202_df <- input_data_eda[input_data_eda$CONTINENT_CODE	=='NN' & input_data_eda$SWIFT_MSG_TYPE=='202',]
NN_202_df <- NN_202_df[,c(3,10,12)]
NN_202_df = NN_202_df %>% distinct()
NN_202_df <- NN_202_df[order(NN_202_df$TRXN_MONTH),]
glimpse(NN_202_df)
dim(NN_202_df)

EU_103_df <- input_data_eda[input_data_eda$CONTINENT_CODE	=='EU' & input_data_eda$SWIFT_MSG_TYPE=='103',]
EU_103_df <- EU_103_df[,c(3,10,12)]
EU_103_df = EU_103_df %>% distinct()
EU_103_df <- EU_103_df[order(EU_103_df$TRXN_MONTH),]
glimpse(EU_103_df)
dim(EU_103_df)

EU_202_df <- input_data_eda[input_data_eda$CONTINENT_CODE	=='EU' & input_data_eda$SWIFT_MSG_TYPE=='202',]
EU_202_df <- EU_202_df[,c(3,10,12)]
EU_202_df = EU_202_df %>% distinct()
EU_202_df <- EU_202_df[order(EU_202_df$TRXN_MONTH),]
glimpse(EU_202_df)
dim(EU_202_df)

AS_103_df <- input_data_eda[input_data_eda$CONTINENT_CODE	=='AS' & input_data_eda$SWIFT_MSG_TYPE=='103',]
AS_103_df <- AS_103_df[,c(3,10,12)]
AS_103_df = AS_103_df %>% distinct()
AS_103_df <- AS_103_df[order(AS_103_df$TRXN_MONTH),]
glimpse(AS_103_df)
dim(AS_103_df)

AS_202_df <- input_data_eda[input_data_eda$CONTINENT_CODE	=='AS' & input_data_eda$SWIFT_MSG_TYPE=='202',]
AS_202_df <- AS_202_df[,c(3,10,12)]
AS_202_df = AS_202_df %>% distinct()
AS_202_df <- AS_202_df[order(AS_202_df$TRXN_MONTH),]
glimpse(AS_202_df)
dim(AS_202_df)
```

## **Bar Plot**

These bar plots reveal the monthly median transaction amounts for each cohort of continent, SWIFT message type and month. 

```{r p4, echo=TRUE, message=FALSE}
library(ggplot2)

options(repr.plot.width = 15, repr.plot.height = 10)

AS_103_df$MONTH_TEXT <- factor(AS_103_df$MONTH_TEXT, levels =AS_103_df$MONTH_TEXT)
ggplot(AS_103_df) +
geom_bar( aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), stat = "identity", fill='#5DADE2', color="#000000") +
geom_line(aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), size = 1.5, color="green", group = 1) +
ggtitle("Asia MT103 Monthly Median Transaction Amount") +
xlab("Month") +
ylab("Median Transaction Amount (Thousands)") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))

AS_202_df$MONTH_TEXT <- factor(AS_202_df$MONTH_TEXT, levels =AS_202_df$MONTH_TEXT)
ggplot(AS_202_df) +
geom_bar( aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), stat = "identity", fill='#5DADE2', color="#000000") +
geom_line(aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), size = 1.5, color="green", group = 1) +
ggtitle("Asia MT202 Monthly Median Transaction Amount") +
xlab("Month") +
ylab("Median Transaction Amount  (Thousands)") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))

EU_103_df$MONTH_TEXT <- factor(EU_103_df$MONTH_TEXT, levels =EU_103_df$MONTH_TEXT)
ggplot(EU_103_df) +
geom_bar( aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), stat = "identity", fill='#5DADE2', color="#000000") +
geom_line(aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), size = 1.5, color="green", group = 1) +
ggtitle("Europe MT103 Monthly Median Transaction Amount") +
xlab("Month") +
ylab("Median Transaction Amount (Thousands)") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))

EU_202_df$MONTH_TEXT <- factor(EU_202_df$MONTH_TEXT, levels =EU_202_df$MONTH_TEXT)
ggplot(EU_202_df) +
geom_bar( aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), stat = "identity", fill='#5DADE2', color="#000000") +
geom_line(aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), size = 1.5, color="green", group = 1) +
ggtitle("Europe MT202 Monthly Median Transaction Amount") +
xlab("Month") +
ylab("Median Transaction Amount (Thousands)") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))

NN_103_df$MONTH_TEXT <- factor(NN_103_df$MONTH_TEXT, levels =NN_103_df$MONTH_TEXT)
ggplot(NN_103_df) +
geom_bar( aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), stat = "identity", fill='#5DADE2', color="#000000") +
geom_line(aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), size = 1.5, color="green", group = 1) +
ggtitle("North America MT103 Monthly Median Transaction Amount") +
xlab("Month") +
ylab("Median Transaction Amount (Thousands)") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))

NN_202_df$MONTH_TEXT <- factor(NN_202_df$MONTH_TEXT, levels =NN_202_df$MONTH_TEXT)
ggplot(NN_202_df) +
geom_bar( aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), stat = "identity", fill='#5DADE2', color="#000000") +
geom_line(aes(x = MONTH_TEXT, y = AVG_TRXN_AMT), size = 1.5, color="green", group = 1) +
ggtitle("North America MT202 Monthly Median Transaction Amount") +
xlab("Month") +
ylab("Median Transaction Amount (Thousands)") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))

```

## **Is the data normally distributed?**

```{r p5, echo=TRUE, message=FALSE}
library(ggplot2)

input_data_eda <- as.data.frame(lapply(input_data_eda, function(x) if(is.numeric(x) && is.na(x)){
  mean(x, na.rm = TRUE)
} else { if(is.character(x) && is.na(x)){x = "NA"} else x }
))

input_data_eda$TRANSACTION_AMOUNT <- input_data_eda$TRANSACTION_AMOUNT/1000000

# hist(input_data_eda$TRANSACTION_AMOUNT, main = "Transaction Amount Frequency Distribution", xlab="Transaction Amount Range (Thousands)")

options(repr.plot.width = 15, repr.plot.height = 10)
ggplot(data = data.frame(input_data_eda$TRANSACTION_AMOUNT), aes(x=input_data_eda$TRANSACTION_AMOUNT)) + geom_histogram(alpha=0.6, bin =50, fill='#5DADE2', color="#000000") + 
xlim(1,18000) + # Removing extreme outlier transaction amount of 0.02
coord_cartesian(ylim=c(0,25000)) +
ggtitle("Transaction Amount Frequency Distribution") +
xlab("Transaction Amount Range (Millions)") +
ylab("Frequency") +
theme(axis.text=element_text(size=12),
      axis.title = element_text(size=12),
      plot.title = element_text(hjust = 0.5, size=15,face="bold"))
```

## **Pie Chart**

North America has the most number of wire tranfers.

```{r p3, echo=TRUE, message=FALSE}

  library(plotly)

  input_data_eda$pie_count = 1

  input_data_eda$CONTINENT_NAME <-  as.factor(input_data_eda$CONTINENT_NAME)

  plot_ly(input_data_eda,
          labels = ~CONTINENT_NAME,
          values = ~pie_count,
          type = 'pie',
          textposition = 'inside',
          textinfo = 'label+percent',
          insidetextfont = list(color = '#FFFFFF'),
          marker = list(colors = colors,line = list(color = '#FFFFFF', width = 2)),
          showlegend = TRUE)  %>%
  layout(title='<b>Transaction % by Continent</b>',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
```

## **Process Runtime**

```{r, echo=TRUE, message=FALSE}
end_time <- Sys.time()
end_time - start_time
```

